---
title: "Downloading Data from Source"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    eval: false
---

```{r}
#| label: setup
#| eval: true
library(panelaR)
```

::: {.callout-important}
## None of the code in this article is executed

Every chunk below is shown for illustration only (`eval: false`). A full run of
these downloaders pulls several years of monthly microdata from three national
statistics offices and takes hours, so the article is rendered without running
it. Copy the chunks into your own session when you actually want the data.
:::

This article walks through
[`scripts/downloading_data_from_source.R`](https://github.com/ctoruno/panelaR/blob/main/scripts/downloading_data_from_source.R),
the script we use to assemble the raw microdata that `panelaR` works on. It has
two jobs, and both matter:

1. **Download** the published microdata for each survey straight from the
   statistical office that produces it (INEC, INE, INEGI).
2. **Flatten** the resulting mess of per-month/per-quarter folders into one
   directory of CSV files per survey, which is the layout
   [`read_data()`](../reference/read_data.html) expects.

Throughout, the destination root is taken from an environment variable so the
script contains no machine-specific paths. Set `PATH_TO_DATA` in your
`.Renviron` to wherever you keep the data.

```{r}
#| label: setup-path
library(panelaR)
path_to_data <- Sys.getenv("PATH_TO_DATA")
```

## Downloading ECU-ENEMDU resources

`download_source_enemdu()` walks every month from `start_year` (2021 by
default, the first year of the post-redesign series) up to the current month,
downloads the CSV and/or SPSS package for each, extracts it into a `YYYY_MM`
subfolder and deletes the ZIP. Months that are already extracted are skipped,
so the function is safe to re-run as new months are published.

```{r}
#| label: download-enemdu
download_source_enemdu(
  file.path(
    path_to_data,
    "ECU-ENEMDU", "source-raw"
  )
)
```

### Why downloading straight from source is fragile here

INEC does not publish ENEMDU behind a stable, predictable set of URLs. The
institution is not consistent in either its **file naming** or its **filing
structure**, and both drift over time:

- The month folder separator changed mid-series, from `Mes-Año` (hyphen, used
  through mid-2022) to `Mes_Año` (underscore, used since).
- Some months carry a `_RECALCULADO` suffix in the file name and some do not.
- The CSV ("Datos Abiertos") and SPSS packages follow two different naming
  templates and are not always published in the same month or at the same time.

`download_source_enemdu()` absorbs this by building every plausible URL for a
given month/format and probing them with a `HEAD` request until one resolves.
That handles most of the drift, but it cannot invent a file that was never
published under any of those patterns — and a handful of months genuinely are
not there.

### The months that must be downloaded by hand

```{r}
#| label: enemdu-manual-gaps
# From Jan-2021 to May-2021, data is not found and has to be manually downloaded from website: https://aplicaciones3.ecuadorencifras.gob.ec/BIINEC-war/
# From Jun-2022 to Jul-2022, data is not found and has to be manually downloaded from website: https://aplicaciones3.ecuadorencifras.gob.ec/BIINEC-war/
# Data for Dec-2022 (CSV) is not found and has to be manually downloaded from website: https://aplicaciones3.ecuadorencifras.gob.ec/BIINEC-war/
# Data for Jul-2023 (SPSS) is not found and has to be manually downloaded from website: https://aplicaciones3.ecuadorencifras.gob.ec/BIINEC-war/
```

**January 2021 to May 2021** is the *RECALCULADO* case, and it is worth
understanding rather than just working around. Those months belong to the
recalculated ENEMDU series: after the 2021 redesign, INEC reprocessed the
Sep-2020–May-2021 files and republished them as `..._RECALCULADO` packages.
The recalculated versions are the ones you want — they are the ones consistent
with the post-redesign series — but they are not exposed under the regular
monthly URL pattern that every other month follows, so no combination of
separator and suffix resolves for them. They have to be pulled from INEC's
BIINEC data portal instead.

The remaining gaps (Jun–Jul 2022, the Dec-2022 CSV, the Jul-2023 SPSS) are
plain publishing omissions from the web directory: the same data is available
in BIINEC, it simply is not filed where the rest of the series lives.

::: {.callout-warning}
## This manual step is required, not optional

The flattening step below reads from **both** `source-raw/` (what the
downloader fetched) and `manual_download/` (what you fetched by hand). If you
skip the manual download, those months are simply missing from the final
dataset, and any panel you build across them will have holes.
:::

So, before going further:

1. Open [INEC's BIINEC portal](https://aplicaciones3.ecuadorencifras.gob.ec/BIINEC-war/).
2. Download the packages for the months listed above.
3. Unpack them anywhere under
   `<PATH_TO_DATA>/ECU-ENEMDU/manual_download/` — the layout inside that folder
   does not matter, because the next step searches it recursively.

### Flattening ENEMDU into a single directory

Downloading is not the end of it. `download_source_enemdu()` leaves you with one
folder per month, and your manual downloads sit in a separate tree entirely.
[`read_data()`](../reference/read_data.html) does not walk directories — it
lists the CSV files in **one** directory, non-recursively, and infers each
file's survey from its name. So the last step is to collect every CSV from both
trees into a single flat `source/` folder.

```{r}
#| label: flatten-enemdu
dest_folder <- file.path(
  path_to_data, "ECU-ENEMDU", "source"
)
dir.create(dest_folder, showWarnings = FALSE)

enemdu_csv_files <- list.files(
  c(
    file.path(path_to_data, "ECU-ENEMDU", "source-raw"),
    file.path(path_to_data, "ECU-ENEMDU", "manual_download")
  ),
  pattern = "\\.csv$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

for (source_file in enemdu_csv_files){
  file.copy(from = source_file, to = dest_folder, overwrite = FALSE)
  message(
    glue::glue("[info] File: {source_file} copied!!")
  )
}
```

Note `overwrite = FALSE`: if a month exists in both trees, whatever landed
first wins and the copy is a no-op. This is what makes the step re-runnable —
run it again after downloading a newly published month and only the new files
are copied.

The result, `<PATH_TO_DATA>/ECU-ENEMDU/source/`, is what you hand to
`read_data()`.

## Downloading CHL-ENE resources

Chile's INE is the easy case. Its files live under a stable URL pattern
(`ene-YYYY-MM-MMM.csv`, one per rolling moving quarter) and are published as
plain CSVs rather than ZIPs — so there is no extraction, no per-period folder,
and consequently nothing to flatten. `download_source_ene()` writes the files
straight into `output_dir` in the layout `read_data()` already wants.

```{r}
#| label: download-ene
download_source_ene(
  file.path(
    path_to_data, "CHL-ENE", "source-raw"
  )
)
```

Files that are not yet published (future months of the current year) are
skipped after a `HEAD` check, and files already on disk are left alone unless
you pass `overwrite = TRUE`.

## Downloading MEX-ENOE resources

INEGI publishes ENOE quarterly, as ZIP archives. Like INEC, it has changed its
naming convention over time — three schemes so far
(`2019trim4_csv.zip`, `enoe_n_2021_trim3_csv.zip`, `enoe_2024_trim1_csv.zip`) —
and `download_source_enoe()` resolves the right one per period automatically.
Each archive is extracted into a per-period subfolder of `output_dir`.

```{r}
#| label: download-enoe
download_source_enoe(
  file.path(
    path_to_data, "MEX-ENOE", "source-raw"
  )
)
```

By default it downloads from 2023 onward (the start of the current ENOE
design) through the current year, in both CSV and DTA. You can target specific
quarters instead with `periods`, e.g.
`download_source_enoe(dir, periods = c("2023_01", "2024_04"))`.

Because the archives extract into per-period folders, ENOE needs the same
flattening step as ENEMDU — the only difference being that there is no manual
download tree to merge in:

```{r}
#| label: flatten-enoe
dest_folder <- file.path(
  path_to_data, "MEX-ENOE", "source"
)
dir.create(dest_folder, showWarnings = FALSE)

enoe_csv_files <- list.files(
  c(
    file.path(path_to_data, "MEX-ENOE", "source-raw")
  ),
  pattern = "\\.csv$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

for (source_file in enoe_csv_files){
  file.copy(from = source_file, to = dest_folder, overwrite = FALSE)
  message(
    glue::glue("[info] File: {source_file} copied!!")
  )
}
```

## Where you end up

Once the script has run, you have one flat directory per survey:

| Survey | Directory | Produced by |
|---|---|---|
| ENEMDU (Ecuador) | `<PATH_TO_DATA>/ECU-ENEMDU/source/` | downloader + manual BIINEC downloads, flattened |
| ENE (Chile) | `<PATH_TO_DATA>/CHL-ENE/source-raw/` | downloader (already flat) |
| ENOE (Mexico) | `<PATH_TO_DATA>/MEX-ENOE/source/` | downloader, flattened |

Each of these can be passed directly to `read_data()`:

```{r}
#| label: read-data
enemdu <- read_data(
  path_to_data = file.path(path_to_data, "ECU-ENEMDU", "source")
)
```
